Exploitation Techniques and Mitigations

### 2016-02-18 docs/paper.tex
Paper: Minor changes regarding content, references and style

@@

Physical memory is managed by the kernel through the use of a Memory Management
Unit (MMU). Each process' address space is virtualized and memory operations
are translated on-the-fly by the MMU. Physical memory is segmented into
\textit{pages} (typically \SI{4}{\kibi\byte} in size) and each page can be
\emph{mapped into} the virtual address space of one or more (shared page)
processes.~\cite[pp.~400]{unix_interals}

The main parts located inside the (virtual) address space of a process are the
executable itself with its \texttt{.text} and \texttt{.data} section, the heap
(used for dynamic data), the stack (used for local variables and function
calling) and libraries.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.4\textwidth]{gfx/stack_smash.png}
    \caption{Stack frame containing a buffer}
    \label{fig:stack_frame}
\end{figure}

@@

@@

After injecting the payload we get a few lines of garbage and receive a prompt
by hitting return a few times. You can enter commands and receive output like
usual.

@@

The first mitigation technique discussed, DEP, is also known under the term
\textit{write XOR execute (\texttt{w\^{}x})} and it will prevent us from
executing injected code like we did in the shell code example previously. This
happens by attaching an execute flag to each page (in addition to the read /
write flags). If the instruction pointer points to a page without the execute
flag set, a segmentation fault will be triggered. The only pages flagged for
execution are the ones that belong either to the \texttt{.text} segment or to
used libraries (by default) since they will contain the program code. The stack
is (of course) not executable by default. Therefore our shell code example
would simply segfault.

The execution flag is enforced by hardware on modern platforms. But for CPUs
that lack such hardware support, software-enforced DEP provides limited
protection.~\cite{wiki:dep}

@@

\begin{listing}[h!]
    \codefile{../src/shell_code/vuln_readelf.txt}
    \caption{\texttt{readelf} output with and without \texttt{-z execstack}}
    \label{src:dep_readelf}
\end{listing}

@@

The first example shown in the \cref{sec:buffer_overflow} already illustrated
the power that comes along with controlling the return address. It enabled us
to jump to a completely different function upon execution. The target binary
may not contain a function doing exactly what an attacker wants to do. But by
controlling the return address one can build a, so called, \textit{ROP chain}
to execute (more or less) arbitrary code, which is made out of
\textit{gadgets}.~\cite{wiki:rop,buchanan2008good,shacham2007}

@@

The final \texttt{ret} statement of a gadget is required for chaining them
together. We not only use the buffer overflow to control \emph{one}, but
\emph{multiple} return addresses. Because of this our payload will be composed
of a list of return addresses (the start of each gadgets) interleaved by some
padding. Each provided address will be consumed one by one at the end of each
gadgets to \emph{jump} to the next one. The payload may still contain data like
\texttt{"/bin//sh"} if necessary.

@@

There are algorithms and tools available which simplify the process of finding
such gadgets (and even whole chains) but they go beyond the scope of this
writing.~\cite{shacham2007}

@@

This mitigation technique was introduced to render ROP (and ret2libc) void. The
idea behind it is quite simple, and the name gives it away already. Memory
layout is randomize so an attacker cannot reliably use ROP. An attacker will
not be able to copy the exact setup of a target machine by only knowing which
binary (and libraries) is used.

@@

We can directly see one flaw in this setup --- not all sections of the
\texttt{cat} binary start at random locations. Especially the \texttt{.text}
always starts at the same position. This happens because \texttt{cat} itself
was not compiled as a Position Independent Executable (PIE). Since this is
actually the default of \texttt{gcc}, most programs' \texttt{.text} segment
will always start at the same location. One could pass the corresponding flag
(\texttt{-pie}) to the compiler to prevent this, so ASLR would be able to
randomize these segments too, but you'd have to compile every relevant package
again instead of using the distribution vendor's pre-compiled binary.

@@

Our previous exploit can be adapted as follows. First, manage to leak a pointer
somehow, which enables you to calculate the address offset introduced by ASLR.
Augment your ROP chain to take the offset into account. Run the exploit. Since
this is rather simple and we already gave an example how to calculate the
offset, we would like to leave this as an exercise for the reader.

@@

CFI is a big topic and, like other topics already mentioned, goes beyond the
scope of this write-up. The first pointer we hand the reader aims at the
corresponding
section\footnote{\url{http://clang.llvm.org/docs/ControlFlowIntegrity.html}} of
the Clang documentation, but we recommend checking out the related research
paper \cite{ms_cfi} for more information. A more accessible and recent way to
this topic may be the
talk\footnote{\url{https://www.youtube.com/watch?v=FA0VK7s5tSQ}} \textit{New
memory corruption attacks: why can't we have nice things?} given by gannimo
(Mathias Payer).

@@

With this mechanism set up, one can easily see that it is no longer possible to
chain \emph{arbitrary} gadgets together to pull off ROP. But we still control
the return address and can jump to different locations as long as we stick to
the CFG. Each transition from one node to another has to be valid, while the
overall path taking by our ROP chain may do things never intended by the
program's author. This concept is known as control-flow bending.~
\cite{cf_bending}

@@

Contrary, breaking ASLR by brute force gets much harder since the address space
is \emph{much} bigger which yields more entropy for the randomization. Similar
techniques like \textit{heap spraying} are basically useless, but we can still
resort to the info leak. Breaking a canary via brute force gets only a little
bit harder, but this has already been shown in an example.

>>>>

\newcommand{\redbox}[1]{%
    \colorlet{currentcolor}{.}%
    {\color{red}%
    \fbox{\color{currentcolor}#1}}%
}

@@

Physical memory is managed by the kernel through the use of a Memory Management
Unit (MMU). Each process' address space is virtualized and memory operations
are translated on-the-fly by the MMU. Physical memory is segmented into
\textit{pages} (typically \SI{4}{\kibi\byte} in size) and each page can be
\emph{mapped into} the virtual address space of one or more (shared page)
processes~\cite[pp.~400]{unix_interals}.

The main parts located inside the (virtual) address space of a process are the
executable itself with its \texttt{.text} and \texttt{.data} segment, the heap
(used for dynamic data), the stack (used for local variables and function
calling) and libraries.

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=0.4\textwidth]{gfx/stack_smash.png}
    \caption{Stack frame containing a buffer~\cite{wiki:buffer_overflow}}
    \label{fig:stack_frame}
\end{figure}

@@

Generally functions handling strings will terminate upon reading a
\texttt{NULL} byte. Offensive Security's \textit{Metasploit Fundamentals} has a
dedicated
section\footnote{\url{https://www.offensive-security.com/metasploit-unleashed/generating-payloads/}
on February 2016} about generating payloads (utilizing Metasploit) without
unwanted bytes. One could work around this manually by setting the first part
of register prior setting the second part. Another way would be by issuing an
\texttt{XOR} with the register itself as in the first instruction of the
example payload.

@@

After injecting the payload we get a few lines of garbage and receive a prompt
by hitting return a few times. You can enter commands and receive output as
usual.

@@

The first mitigation technique discussed, DEP, is also known under the term
\textit{write XOR execute (\texttt{w\^{}x})} and it will prevent us from
executing injected code as we did in the previous shell code example. This
happens by attaching an execute flag to each page (in addition to the read /
write flags). If the instruction pointer points to a page without the execute
flag set, a segmentation fault will be triggered. The only pages flagged for
execution are the ones that belong either to the \texttt{.text} segment or to
used libraries (by default) since they will contain the program code. The stack
is (of course) not executable by default. Therefore our shell code example
would simply segfault.

The execution flag is enforced by hardware on modern platforms. But for CPUs
that lack such hardware support, software-enforced DEP provides limited
protection~\cite{wiki:dep}.

@@

\begin{listing}[h!]
    \codefile[escapeinside=\%\%]{../src/shell_code/vuln_readelf.txt}
    \caption{\texttt{readelf} output with and without \texttt{-z execstack}}
    \label{src:dep_readelf}
\end{listing}

@@

The first example shown in \cref{sec:buffer_overflow} already illustrated the
power that comes along with controlling the return address. It enabled us to
jump to a completely different function upon execution. The target binary may
not contain a function doing exactly what an attacker wants to do. But by
controlling the return address one can build a, so called, \textit{ROP chain}
to execute (more or less) arbitrary code, which is made out of
\textit{gadgets}~\cite{wiki:rop,buchanan2008good,shacham2007}.

@@

The final \texttt{ret} statement of a gadget is required for chaining them
together. We not only use the buffer overflow to control \emph{one}, but
\emph{multiple} return addresses. Because of this our payload will be composed
of a list of return addresses (the start of each gadget) interleaved by some
padding. Each provided address will be consumed one by one at the end of each
gadget to \emph{jump} to the next one. The payload may still contain data such
as \texttt{"/bin//sh"} if necessary.

@@

Various tools simplifying the process of finding gadgets (and even whole
chains) already exist, for instance
\texttt{ROPgadget.py}\footnote{\url{https://github.com/JonathanSalwan/ROPgadget}},
but they go beyond the scope of this writing~\cite{shacham2007}.

@@

This mitigation technique was introduced to render ROP (and ret2libc) void. The
idea behind it is quite simple, and the name gives it away already. Memory
layout is randomized so an attacker cannot reliably use ROP. An attacker will
not be able to copy the exact setup of a target machine by only knowing which
binary (and libraries) is used.

@@

We can directly see one flaw in this setup --- not all sections of the
\texttt{cat} binary start at random locations. Especially the \texttt{.text}
always starts at the same position. This happens because \texttt{cat} itself
was not compiled as a Position Independent Executable (PIE). Since this is
actually the default of \texttt{gcc}, most programs' \texttt{.text} segment
will always start at the same location. One could pass the corresponding flag
(\texttt{-pie}) to the compiler to prevent this, so ASLR would be able to
randomize these segments too, but one would have to compile every relevant
package again instead of using the distribution vendor's pre-compiled binary.

@@

Our previous exploit can be adapted as follows. First, manage to leak a pointer
somehow, which enables you to calculate the address offset introduced by ASLR.
Augment your ROP chain to take the offset into account. Run the exploit. Since
this is rather simple and we already gave an example how to calculate the
offset, we leave this as an exercise for the reader.

@@

CFI is a big topic and, similar to other topics already mentioned, goes beyond
the scope of this write-up. The first pointer we hand the reader aims at the
corresponding
section\footnote{\url{http://clang.llvm.org/docs/ControlFlowIntegrity.html}} of
the Clang documentation, but we recommend checking out the related research
paper \cite{ms_cfi} for more information. A more accessible and recent way to
this topic may be the
talk\footnote{\url{https://www.youtube.com/watch?v=FA0VK7s5tSQ}} \textit{New
memory corruption attacks: why can't we have nice things?} given by gannimo
(Mathias Payer).

@@

With this mechanism set up, one can easily see that it is no longer possible to
chain \emph{arbitrary} gadgets together to pull off ROP. But we still control
the return address and can jump to different locations as long as we stick to
the CFG. Each transition from one node to another has to be valid, while the
overall path taking by our ROP chain may do things never intended by the
program's author. This concept is known as control-flow bending~
\cite{cf_bending}.

@@

Contrary, breaking ASLR by brute force gets much harder since the address space
is \emph{much} bigger which yields more entropy for the randomization. Similar
techniques such as \textit{heap spraying} are basically useless, but we can
still resort to the info leak. Breaking a canary via brute force gets only a
little bit harder, but this has already been shown in an example.
